Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-27 paper/paper.tex
more futzing with figure, we have room for more columns if we want
intro

\section{Introduction}
In this paper, we present a \emph{short} and \emph{mechanized} proof of
logical consistency for \lang{}, a dependent type theory with a full
predicative universe hierarchy, large eliminations, an intensional identity
type, a boolean base type, and dependent elimination forms.

Our goal with this work is to demonstrate how the proof technique of
\emph{syntactic logical relations} can be applied to reason about dependent
type theories.  Logical relations are a powerful proof technique for reasoning
about lambda calculi and their operational semantics. They have been used to
show diverse properties such as strong
normalization~\citep{girard1989proofs,
  geuvers1994short}, contextual equivalence~\citep{constable1986implementing},
representation independence~\citep{pitts1998existential}, noninterference~\citep{bowman2015noninterference}, compiler correctness~\citep{benton2009biorthogonality,perconti2014compiler}, and the
decidability of conversion algorithms~\citep{harper2005equivalence,Abel12,abel2013normalization}. \scw{% add citations?
  emphasize that we are working
with syntactic logical relations.}

Tutorial material on logical relations~\citep{skorstengaard2019introduction,
  harpertait, harperkripke, pierce2002types,
  pierce2004advanced,harper2016practical} is primarily focused on systems
with simple or polymorphic types. In that context, syntactic logical relations
can be defined as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation. However, neither of these techniques can be used to define a
logical relation in the context of a predicative dependent type theory, so a
young researcher might be excused for thinking that proofs that use logical
relations are not applicable for such languages.

@@

Thus, our paper provides a gentle and accessible introduction to a powerful
technique for dependent type theories. To promote the use of machine-assisted
reasoning, our development is accompanied by a short mechanized proof script,
of less than 1000 lines of code, developed using the Coq/Rocq proof
assistant~\cite{coq}.

We have streamlined our proof through a number of means: the careful selection
of the features that we include in the object type system and the results
that we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, but we do include propositional
equality and booleans to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we prove logical consistency, which states that empty types are not inhabited in an empty context, and demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal forms for well-typed open \emph{and}
closed terms. We include a full predicative universe hierarchy and
type-level computation to demonstrate the logical strength of the
approach. % \scw{Say that we don't have impredicativity, but neither does anyone
  % else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

The result of our work is a development that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set-theoretic notation and terminology so that
the material is accessible to readers with a general mathematical background.
That said, our explanations do not stray too far away from our proof scripts.
We link each lemma directly to its counterpart in the proof script,
anticipating that readers may wish to reference and validate our formalism.
The typeset proofs purposefully follow the
mechanized proofs closely while avoiding, as much as possible, artifacts specific to Coq.

Not only does this connection aid readers that wish to, like us, adopt proof
assistants for their day-to-day use, we also find that this precision is
important for conveying the proof technique itself. Unlike properties that are
derivable through syntactic means, proofs by logical relation make demands on
the strength of the logic in which they are expressed. An informal proof that
attempts to be agnostic or ambiguous about the underlying
metatheory % may struggle to even
% convey the validity of their definition and
requires substantial
effort from the reader to even understand whether it is theoretically
possible to encode such proofs in a given ambient logic, such as those found
in modern proof assistants.

% The close correspondence between a typeset proof and a mechanized
% proof comes at the potential risk that the artifacts specific to
% theorem provers may become visible in the presented proof.
%However, for proofs by logical relation for
%dependent types, we find our approach advantageous for a few reasons.
Furthermore, while a proof assistant may reject perfectly valid definitions,
it helps us keep our definition precise and unambiguous. A precise definition
is crucial, as the reasoning principles we are allowed to use are directly tied
to the definition, which is sometimes obscured by concise notations that
diverge too far from a formal proof.

@@

\yl{new overview starts here}
More concretely, our paper makes the following contributions. 
\begin{itemize}
\item In Section~\ref{sec:spec},
we present the type system of \lang{},
a dependently typed system with an untyped conversion rule in the
style of Pure Type Systems~\citep{barendregt1991introduction}. The
\lang{} language supports many common features of a dependent
type theory: a countable universe hierarchy, a boolean base
type, an intensional identity type, and support for large elimination.
\item In Section~\ref{sec:logreldep}, we formulate logical consistency for \lang{}, the property of
interest, to motivate a logical relation, and proceed to define it as
an inductive relation which we later prove to be a partial
function. Based on the definition of the logical relation, we define
a notion of semantic typing and prove the fundamental theorem, from
which logical consistency follows as a corollary
(Section~\ref{sec:logrelproof}). Thanks to the good coverage of
features in our object language, our proof showcases the
special treatment required to model those features, thus making
our proof applicable to a broad range of type systems.
\item We strengthen our logical relation to prove the existence of
  $\beta$-normal forms (Section~\ref{sec:extension}) and
  $\beta\eta$-normal forms (Section~\ref{sec:betaeta}) for well-typed
  open \emph{and} closed terms. The modifications made to the logical
  relation are small and closely mirror an analogous extension for a
  simply typed language. We use these extensions to demonstrate the idea
  that once we have established the base techniques, we can port ideas
  that have been studied in the context of a simpler system to the
  dependently typed setting.
\item We mechanize all our proofs using the Coq proof assistant, with
  963 lines of code for the consistency proof and a moderate increase
  to 1357 lines of code for the normalization proof.
  We discuss the specifics
  related to our choice of Coq as our metatheory, including our use of
  off-the-shelf semantic engineering infrastructure and automation
  tools, in Section~\ref{sec:logrelmech}.
  Our proof scripts, with comments, are available to reviewers as
  supplementary materials. We use Coq's custom notations to ease the
  effort of matching the pen and paper representation to our formalism.
 \yl{I want to say something
  about the benefits of using existing libraries such as Autosubst 2
  and CoqHammer rather than rolling your own undocumented libraries that nobody
  want to (or can) reuse.}
\scw{I'm not sure that this is the best place for this discussion.}
Our development relies on off-the-shelf tools...
\item We discuss in Section~\ref{sec:relatedwork} how our proof
relates to existing proofs by logical relations and other proof
techniques for proving consistency and normalization. In
Section~\ref{sec:discuss}, we give an in-depth explanation on how
various design decisions (e.g. the choice of untyped equality) affect
the size of our proof and its extensibility to features such as
type-directed conversion.
\end{itemize}
We hope our success at creating a short and mechanized proof for a
relatively feature-complete language will encourage future researchers
to leverage the tool of logical relations in mechanized
proofs for dependent types.

\yl{new overview ends here}

The key contributions of each section of our paper are as follows.
\scw{I'm still not happy with this overview / contributions list. }

In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type theory of
interest, reminiscent of Martin-Löf style predicative type systems. A key
design choice that impacts our proofs is the use of an untyped conversion
rule, inspired by Pure Type Systems~\citep{barendregt1991introduction}, and
specified by parallel
reduction~\citep{takahashi-parallel-reduction, barendregt:lambda-calculi-with-types} % to normal form~\citet{barendregt}
.
Because we believe that a good coverage of features makes our proof applicable
to a broad range of type systems, \lang{} incorporates some of the most common
features of dependent type theory, including an intensional identity type, a
boolean type, large elimination, dependent elimination, and a full universe
hierarchy. While our use of untyped conversion is nonstandard for MLTT, we
discuss how our proof can be adjusted for systems with a typed, judgmental
equality.

In Section~\ref{sec:logreldep}, we formulate the consistency property and
motivate where a proof by logical relations is required. The most important
contributions of this section are the definition of the logical relation as an
inductively defined relation and demonstrating how to reason about this
definition. We also show that the logical relation respects evaluation
(referred to as the irrelevance property in \citet{martin-lof-a-la-coq}), and
prove that the logical relation is a partial function: that is, each type can
have at most one unique interpretation.

In Section~\ref{sec:logrelproof}, we show that our definitions can be the
subject of well-established reasoning techniques for logical relations. In
particular, we define a notion of semantic typing in terms of our logical
relation and prove the fundamental theorem, which states that syntactic typing
implies semantic typing. Consistency follows from this result as a corollary.

In Section~\ref{sec:extension}, we extend the definition of the logical relation
to include terms with free variables, and prove that every well-typed term
has a $\beta$-normal form. After laying the foundation for handling type-level
computation in Section~\ref{sec:logreldep}, we show how this extension requires mechanical
changes that directly mirror an analogous extension for a simply typed
language. We use this section to demonstrate the idea that once we have
established the base techniques, we can port ideas that have been studied in
the context of a simpler system to the dependently typed setting.  We believe
it is possible to extend our logical relation to a Kripke-style and relational
model in a similar fashion.

In Section~\ref{sec:betaeta}, we show how to incorporate the function $\eta$ law
into the conversion rule for \lang{}.

In Section~\ref{sec:logrelmech}, we discuss the details of our
mechanization, including our use of existing libraries such as
Autosubst 2~\citep{autosubst2} for handling bindings and
CoqHammer~\citep{czajka2018hammer} for general-purpose automation.

In Section~\ref{sec:relatedwork}, we discuss how our proof relates to
existing proofs by logical relations and other proof techniques for
proving consistency and normalization.

In Section~\ref{sec:discuss}, we explain the various factors that contribute
to our short proof.

We hope our success at creating a short and mechanized proof for a
relatively feature-complete language will encourage future researchers
to leverage the tool of logical relations in mechanized
proofs for dependent types.

\scw{TODO: this introduction is still not done. At this point, I think the
  first part has gotten a little long. We could say what we mean more
  concisely. We also need to work on the overview at the end of this section
  --- it need to be more descriptive about contributions and less of a table
  of contents.}

% % For
% % example, \citet{geuvers1994short} requires impredicativity to encode
% % Calculus of Constructions, an impredicative dependent type theory, but
% % also relies on the classification property to define the logical
% % relation over well-formed types and kinds. Their simple
% % presentation of the logical relation as an inductively defined
% % function over types, reminiscent of logical relations for simply typed
% % languages,

@@

\section{Logical Relation}
% \scw{This section pre-supposes that we want to define a logical relation,
% but doesn't precisely state what we want to use it to prove, and why a logical
% relation is suitable. (And why the property you want to prove is difficult to
% show!) Should add more motivation here.}
\scw{We need to explicitly point out that the key ideas of this paper are
  discussed, here, in this section.
  We need to explicitly remark on why logical relations are difficult to
  define for dependent type theory and explain why this setting is more
  difficult than with simple types (STLC) or with polymorphic types (System F).
  \begin{itemize}
  \item Dependent function types (result type mentions argument value, so can't define
    relation by recursion over type structure)
  \item Definitional equality (not all types look like types)
  \item Identity types(?)
  \end{itemize}
  Should we be more explicit in our comparison with Girard's trick for polymorphic type?
  There, the definition stays recursive because it doesn't substitute for the variables
  in the function types. But that approach is not available in this setting, because not
  all quantified things are types. And we might need that information to interpret, say,
  identity types in the right way.
}
\yl{I don't think identity types and dependent function types should
  be listed as reasons why the logrel can't be a recursive
  function. System DE has both dependent function types and identity
  types but the logical fragment doesn't have type-level computation
  so it can be defined structurally by specifying some kind of size metric
}
\scw{ We also need to explicitly point out that our logical relation is untyped.
  This has two benefits: it allows semantic typing to be meaningful independent from
  syntactic typing (cite Derek, forward reference to next section) and it avoids
  significant bookkeeping, especially in the case of Kripke logical relations (we need to define
  what these are).
  Is there a cost to an untyped relation?
}

@@

\subsection{Definition of the Logical Relation}
\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
% \scw{Many introductory texts define the relation as a recursive function
% over type structure, or step-indices. You use an inductive relation instead, why?}
%\yl{resolved}
% \scw{Can we view this inductive relation as the graph of the partial function
% that is defined recursively over types?} \yl{reolsved}
% \scw{What is this form extensible too? impredicative quantification? recursive
% types? } \yl{resolved? impredicativity is hard}
% \scw{Is it worth observing here that this definition is not over sets of typed
% terms. That it characterizes all terms that look like booleans (i.e. evaluate to
% true or false) or all terms that look like proofs (i.e. evaluate to refl). The
% fact that there is no connection between p and a and b in the I-Eq case is strange
% looking. Need to explain.  }
% \yl{resolved}
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative and we will explain the reason shortly.

@@

% \begin{minipage}{0.8\textwidth}
  \begin{tabular}{| l |  l  | l | l | l | l| }
    \hline
%      & \header{Universes} & \header{Inductives} & \header{Conversion}
%      & \header{Large Elim} & Main results \\
      & \header{Univ} & \header{Ind} & \header{Conv}
      & \header{Large E} & Main results \\

    \hline
    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓}
    & Consistency and normalization \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗} & Consistency \\
    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & Correctness of NBE \\

    % \hline
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} & Decidability of conversion \\
    % \hline
    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} & Decidibility of type checking \\
    % \hline
    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & Consistency \\
    \hline
    % CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
    % CoreSpec & \\
    % PCUIC & \\
    %   & \lang{} & \citet{anand2014towards} &
    %                                          \citet{barras2012semantical}
    % & \citet{martin-lof-a-la-coq}  \\
    % \hline
    % Universe & Countable & Countable & Impredicative + Countable & One\\
    % % \hline
    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
    % \hline
    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
    % \hline
    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
  \end{tabular}

  \begin{tabular}{ll}
    \\
  \underline{Univ}erses: & Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
  \underline{Ind}uctives: & Identity types (Id), Natural numbers (Nat), W-types \\
  \underline{Conv}ersion:& Untyped (U), Typed (T),  Extensional (E) \\
  \underline{Large E}liminations:& Included ({\boxedsymbols ✓}), not included ({\boxedsymbols ✗}) \\
  \\
  $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
  NBE-in-Coq& \citet{nbeincoq} \\
  $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
  MLTT-a-la-Coq &\citet{martin-lof-a-la-coq} \\
  Core Nuprl &\citet{anand2014towards} \\
  \end{tabular}

>>>>

\section{Introduction}
This paper presents a \emph{short} and \emph{mechanized} proof of
logical consistency for \lang{}, a dependent type theory with a full
predicative universe hierarchy, large eliminations, an intensional identity
type, a boolean base type, and dependent elimination forms.

Our goal with this work is to demonstrate the application of the proof
technique of \emph{syntactic logical relations} to dependent type theories.
Logical relations are a powerful proof technique, and have been used to show
diverse properties such as strong normalization~\citep{girard1989proofs,
  geuvers1994short}, contextual equivalence~\citep{constable1986implementing},
representation independence~\citep{pitts1998existential},
noninterference~\citep{bowman2015noninterference}, compiler
correctness~\citep{benton2009biorthogonality,perconti2014compiler}, and the
decidability of conversion
algorithms~\citep{harper2005equivalence,Abel12,abel2013normalization}.

However, tutorial material on syntactic logical
relations~\citep{skorstengaard2019introduction, harpertait, harperkripke,
  pierce2002types, pierce2004advanced,harper2016practical} is primarily
focused on systems with simple or polymorphic types. In that context,
syntactic logical relations can be defined as simple recursive functions over
the structure of types, or (in the case of recursive types) defined over the
evaluation steps of the computation. Yet, neither of these techniques can be
used to define a logical relation in the context of a predicative dependent
type theory, so a novice researcher might be excused for thinking that proofs
that use logical relations are not applicable for such languages.

@@

Thus, our paper provides a gentle and accessible introduction to a powerful
technique for dependent type theories. To promote the use of machine-assisted
reasoning, our development is accompanied by a short mechanized proof script,
of less than 1000 lines of code, developed using the Coq proof
assistant~\cite{coq}.

We have streamlined our proof through a number of means: the careful selection
of the features that we include in the object type system and the results that
we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, but we do include propositional
equality and booleans to capture the challenges presented by indexed types and
dependent pattern matching. We do not show the decidability of type checking,
nor do we develop a PER semantics, but we prove logical consistency, which
states that empty types are not inhabited in an empty context, and demonstrate
how our consistency proof can be extended (at a moderate cost of 500 lines of
code) to show the existence of $\beta\eta$-normal forms for well-typed open
\emph{and} closed terms. We include a full predicative universe hierarchy and
type-level computation to demonstrate the logical strength of the approach.

%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

More concretely, our paper makes the following contributions. 
\begin{itemize}
\item In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type
  theory of interest. A key design choice that impacts our proofs is the use
  of an untyped conversion rule, inspired by Pure Type
  Systems~\citep{barendregt1991introduction}, and specified through parallel
  reduction~\citep{takahashi-parallel-reduction,
    barendregt:lambda-calculi-with-types}.
\item In Section~\ref{sec:logreldep}, we formulate logical consistency for
  \lang{}, the property of interest, to motivate a logical relation. We define
  the relation first inductively and then later prove that it is a partial
  function. Based on this definition, we define semantic typing and prove the
  fundamental theorem, from which consistency follows as a corollary
  (Section~\ref{sec:logrelproof}). Thanks to the design of \lang{}, our proof
  showcases the special treatment required to model many of the most common
  features of dependent type theories, thus making our proof applicable to a
  broad range of type systems.
\item We strengthen our logical relation to prove the existence of
  $\beta$-normal forms (Section~\ref{sec:extension}) and $\beta\eta$-normal
  forms (Section~\ref{sec:betaeta}) for well-typed open terms. The
  modifications made to our initial logical relation are small and closely
  mirror the necessary extensions for a simply-typed language. We use this
  part to show that once we have established the base techniques, we can port
  ideas from proofs about simpler languages to the dependently typed setting.
\item We mechanize all our proofs using the Coq proof assistant, with 963
  lines of code for the consistency proof and a moderate increase to 1357
  lines of code for the normalization proof.  We discuss the specifics related
  to our choice of Coq as our metatheory, including our use of off-the-shelf
  semantic engineering infrastructure and automation tools, in
  Section~\ref{sec:logrelmech}.  Our proof scripts, with comments, are
  available to reviewers as supplementary materials.
%  We use Coq's custom notations to ease the
%  effort of matching the pen and paper representation to our formalism.
  \yl{I want to say something about the benefits of using existing libraries
    such as Autosubst 2 and CoqHammer rather than rolling your own
    undocumented libraries that nobody want to (or can) reuse.}\scw{Save that
    for the mechanization section}
% Our development relies on off-the-shelf tools...
\item We compare our work to existing proofs by logical relations and other
  proof techniques for proving consistency and normalization.  We provide an
  overview of this prior work (Section~\ref{sec:relatedwork}) and also give an
  in-depth explanation on how various design decisions affect the size of our
  proof and its extensibility to additional features
  (Section~\ref{sec:discuss}).
\end{itemize}

The result of our work is an artifact that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set-theoretic notation and terminology so
that the material is accessible to readers with a general mathematical
background.  That said, our explanations do not stray too far away from our
proof scripts.  We link each lemma directly to its counterpart in the proof
script, anticipating that readers may wish to reference and validate our
formalism.  The typeset proofs purposefully follow the mechanized proofs
closely while avoiding, as much as possible, artifacts specific to Coq.

Not only does this connection aid readers that wish to, like us, adopt proof
assistants for their day-to-day use, we also find that this precision is
important for conveying the proof technique itself. Unlike properties that are
derivable through syntactic means, proofs by logical relation make demands on
the strength of the logic in which they are expressed. An informal proof that
attempts to be agnostic or ambiguous about the underlying metatheory requires
substantial effort from the reader to even understand whether it is
theoretically possible to encode such proofs in a given ambient logic, such as
those found in modern proof assistants.

@@

@@

\section{Logical Relation}
\label{sec:logreldep}

% \scw{This section pre-supposes that we want to define a logical relation,
% but doesn't precisely state what we want to use it to prove, and why a logical
% relation is suitable. (And why the property you want to prove is difficult to
% show!) Should add more motivation here.}
\scw{We need to explicitly point out that the key ideas of this paper are
  discussed, here, in this section.
  We need to explicitly remark on why logical relations are difficult to
  define for dependent type theory and explain why this setting is more
  difficult than with simple types (STLC) or with polymorphic types (System F).
  \begin{itemize}
  \item Dependent function types (result type mentions argument value, so can't define
    relation by recursion over type structure)
  \item Definitional equality (not all types look like types)
  \item Identity types(?)
  \end{itemize}
  Should we be more explicit in our comparison with Girard's trick for polymorphic type?
  There, the definition stays recursive because it doesn't substitute for the variables
  in the function types. But that approach is not available in this setting, because not
  all quantified things are types. And we might need that information to interpret, say,
  identity types in the right way.
}
\yl{I don't think identity types and dependent function types should
  be listed as reasons why the logrel can't be a recursive
  function. System DE has both dependent function types and identity
  types but the logical fragment doesn't have type-level computation
  so it can be defined structurally by specifying some kind of size metric
}
\scw{ We also need to explicitly point out that our logical relation is untyped.
  This has two benefits: it allows semantic typing to be meaningful independent from
  syntactic typing (cite Derek, forward reference to next section) and it avoids
  significant bookkeeping, especially in the case of Kripke logical relations (we need to define
  what these are).
  Is there a cost to an untyped relation?
}

@@

\subsection{Definition of the Logical Relation}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
% \scw{Many introductory texts define the relation as a recursive function
% over type structure, or step-indices. You use an inductive relation instead, why?}
%\yl{resolved}
% \scw{Can we view this inductive relation as the graph of the partial function
% that is defined recursively over types?} \yl{reolsved}
% \scw{What is this form extensible too? impredicative quantification? recursive
% types? } \yl{resolved? impredicativity is hard}
% \scw{Is it worth observing here that this definition is not over sets of typed
% terms. That it characterizes all terms that look like booleans (i.e. evaluate to
% true or false) or all terms that look like proofs (i.e. evaluate to refl). The
% fact that there is no connection between p and a and b in the I-Eq case is strange
% looking. Need to explain.  }
% \yl{resolved}
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative and we will explain the reason shortly.

@@

% \begin{minipage}{0.8\textwidth}
  \begin{tabular}{| l |  c  | l | c | c | l | }
    \hline
%      & \header{Universes} & \header{Inductives} & \header{Conversion}
%      & \header{Large Elim} & Main results \\
      & \header{Univ} & \header{Ind} & \header{Conv}
      & \header{L E} & Main results \\

    \hline
    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓}
    & Consistency and normalization \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗} & Consistency \\
    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & Consistency \\
    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & Correctness of NBE \\
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} & Decidability of conversion \\
    % \hline
    % \hline
    MLTT-\'a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} & Decidibility of type checking \\
    % \hline
    \hline
    % CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
    % CoreSpec & \\
    % PCUIC & \\
    %   & \lang{} & \citet{anand2014towards} &
    %                                          \citet{barras2012semantical}
    % & \citet{martin-lof-a-la-coq}  \\
    % \hline
    % Universe & Countable & Countable & Impredicative + Countable & One\\
    % % \hline
    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
    % \hline
    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
    % \hline
    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
  \end{tabular}

  \begin{tabular}{ll}
    \\
  \underline{Univ}erses: & Countable ($\mathbb{N}$), Zero (0), One (1) \\
  \underline{Ind}uctives: & Identity types (Id), Natural numbers (Nat), W-types \\
  \underline{Conv}ersion:& Untyped (U), Typed (T),  Extensional (E) \\
  \underline{L}arge \underline{E}liminations:& Included ({\boxedsymbols ✓}), not included ({\boxedsymbols ✗}) \\
  \\
  $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
  Core Nuprl &\citet{anand2014towards} \\
  $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
  NBE-in-Coq& \citet{nbeincoq} \\
  MLTT-\'a-la-Coq &\citet{martin-lof-a-la-coq} \\
  \end{tabular}
